Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-14 agujournaltemplate.tex
Final Text Revision

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are photosynthesizing unicellular microorganisms that form the foundation of the ocean food web. Since they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study investigated how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state antecedent 2004. Interestingly, community diversity declined between 1998 and 2009 independently from changes in the upwelling regime. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

% PARAGRAPH THREE - CHANGES IN CARIACO
Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". But, to our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a mooring station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Analysis of Phytoplankton Community Structure}
    To investigate the structure of the phytoplankton community over time, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. 
    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis. Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

>>>>

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are photosynthesizing unicellular microorganisms that form the foundation of the ocean food web. Since they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study investigated how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state preceding 2004. Interestingly, community diversity declined between 1998 and 2009 independently from changes in the upwelling regime. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biological parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

% PARAGRAPH THREE - CHANGES IN CARIACO
Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global climate change. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period caused a regime change in the physical and biogeochemical system \cite{taylor_ecosystem_2012}. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''. To our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the local ecological responses to environmental variability across decadal scales.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological, chemical and physical data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a mooring station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS (see Table \ref{tab:sup:GenusFuncGroups} in Appendix).

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Analysis of Phytoplankton Community Structure}
    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually. By using binary data and temporal aggregation, we aimed to reduce the influence of extreme values in the cell counts and focus on broad compositional shifts in the community. 
    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination on the monthly counts, depth-integrated to \qty{100}{\meter}, that were square-root transformed using ``metaMDS'' from the ``vegan'' package \cite{oksanen_vegan_2024}. The NMDS analysis is based on the Bray distance matrix with a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the ``envfit'' function from the ``vegan'' package to fit the environmental vectors onto the ordination with 999 permutations. In contrast to the annual clustering, the NMDS ordination uses the individual monthly cell counts and takes abundance of genera into account.
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the ``ggpubr'' package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the ``wilcox.test'' function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis. Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community changes. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$; see Table \ref{tab:DataSources} for a full overview). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift in the community. This shift persists during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.
